Simulations of Ballot Polling Risk-Limiting Audits

### 2023-02-13 providence/sections/appendix.tex
mostly proofs

\begin{lemma}
    \label{lemma:frac_sums_increasing}
    Given a monotone increasing sequence: $\frac{a_1}{b_1}, \frac{a_2}{b_2}, \ldots, \frac{a_n}{b_n}$, for $a_i, b_i > 0$, the sequence:
    $$z_i = \frac{\sum_{j=i}^n a_j}{\sum_{j=i}^n b_j}$$
    is also monotone increasing.
\end{lemma}

@@

\begin{lemma}
\label{lemma:any_ratio_is_sigma_simple}
For $j\ge 1$,
$$\frac{Pr[\bm{K_j}=\bm{k_j} \mid \bm{n_j}, H_a]}{Pr[\bm{K_j}=\bm{k_j} \mid \bm{n_j}, H_0]} = \sigma(k_j, p_a, p_0, n_j).$$
\end{lemma}
\begin{proof}
We induct on the number of rounds.
For $j=1$, we have
$$\frac{Pr[\bm{K_1}=\bm{k_1} \mid \bm{n_1},H_a]}{Pr[\bm{K_1}=\bm{k_1} \mid  \bm{n_1},H_0]} =\frac{Pr[K_1 = k_{1} \mid n_1,H_a]}{Pr[K_1 = k_1 \mid n_1,H_0]} $$$$= \frac{\text{Bin}(k_1,n_1,p_a)}{\text{Bin}(k_1,n_1,p_0)}=\sigma(k_1, p_a, p_0, n_1).$$
Suppose the lemma is true for round $j=m$ with history $\bm{k_m}$.
Observe that
 $$\frac{Pr[\bm{K_{m+1}}=\bm{k_{m+1}} \mid \bm{n_{m+1}},H_a]}{Pr[\bm{K_{m+1}}=\bm{k_{m+1}} \mid \bm{n_{m+1}}, H_0]} $$$$= \frac{ Pr[\bm{K_{m}}=\bm{k_{m}}\mid \bm{n_{m+1}},H_a] \cdot Pr[K_{m+1}'=k_{m+1}'|\bm{k_m},\bm{n_{m+1}},H_a]}{ Pr[\bm{K_{m}}= \bm{k_{m}} \mid  \bm{n_{m+1}},H_0]  \cdot  Pr[K_{m+1}'=k_{m+1}'|\bm{k_m},\bm{n_{m+1}},H_0]  }$$
 $$=\sigma(k_m, p_a, p_0, n_m) \cdot \frac{Pr[K_{m+1}'=k_{m+1}'|\bm{k_m}, \bm{n_{m+1}}, H_a]}{Pr[K_{m+1}'=k_{m+1}'|\bm{k_m},\bm{n_{m+1}},H_0]}$$
 by the induction hypothesis.
Then this is simply equal to
 $$\sigma(k_m, p_a, p_0, n_m)\cdot\frac{\text{Bin}(k_{m+1}',n_{m+1}',p_a)}{\text{Bin}(k_{m+1}',n_{m+1}',p_0)}
 $$$$
 =\frac{p_a^{k_m} (1-p_a)^{n_m-k_m}}{p_0^{k_m} (1-p_0)^{n_m-k_m}} \cdot
 \frac{p_a^{k_{m+1}'} (1-p_a)^{n_{m+1}'-k_{m+1}'}}{p_0^{k_{m+1}'} (1-p_0)^{n_{m+1}'-k_{m+1}'}}
 $$
 $$
 =\sigma(k_{m+1}, p_a, p_0, n_{m+1})
 $$
\end{proof}

\begin{definition} 
\label{def:kmin}
Let $[n_1, \ldots, n_j]$ be the round schedule of an audit that has not stopped by the round $j-1$. Let us define 
\begin{small}
\begin{equation}\label{eq:kMin}
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  =
  \min\left\{k : \omega_j(k, k_{j-1},p_a,p_0,n_j, n_{j-1}) \geq \frac{1}{\alpha}  \right\}.
%  \min\left\{k : \sigma(k_{r-1},p_a,p_0,n_{r-1})\cdot \tau_1(k-k_{r-1},p_a,p_0,n_r-n_{r-1}) \geq \frac{1}{\alpha}  \right\}$
\end{equation}
\end{small}
\end{definition}
As we have seen in Lemma \ref{lemma:minerva2_kmin_exists}, such a value of $k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}$ exists and $k_j \geq k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $ if and only if the result of the audit is Correct, (\textit{i.e.,} the stopping condition in Definition~\ref{def:minervatwo} holds).

The following lemma shows a Markov-like property of \Providence audit (\textit{i.e.,}
for an audit that has not stopped in the first $j-1$ rounds, only cumulative results of the round $j-1$ matter: cumulative sample size $n_{j-1}$ and the number of ballots for the winner $k_{j-1}$).

\textbf{Lemma \ref{lemma:markov}.}
Let $[n_1, \ldots, n_{j-1}, n_j]$ be a round schedule for an execution of  \Providence audit that has not stopped
in any of its first $j-1$ rounds (\textit{i.e.,} for every $i = 1, \ldots, j-1$:
$k_i < k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $), then:

\[ 
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} = k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j} .
\]
% \fpo{this can be used to prove that \Providence is more efficient than \Minerva and \BRAVO}
\begin{proof}
Let $k_{j-1}$ denote the number of ballots drawn for the declared winner up to the round $j-1$ (out of $n_{j-1}$ sampled ballots). The stopping decision for the round $j$ is made as follows:

\[
 k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  = \min\left\{k : \omega_{j}(k, k_{r-1}, p_a, p_0, n_r, n_{r-1}) \geq \frac{1}{\alpha}  \right\} = 
\]
\[
  =  k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j}  
\]

\end{proof}

That is, the stopping condition is equivalent to that of a two round audit with the same cumulative votes for the winner and cumulative round sizes: the first round is of size $n_{j-1}$ and has $k_{j-1}$ votes for the winner, and the second (cumulative) round size is $n_j$ with $k_j$ (cumulative) votes for the winner. Compare this to the similar property for the $\Bravo$ stopping condition.

>>>>

\begin{lemma}
    \label{lemma:frac_sums_increasing}
    Given a monotone increasing sequence: $\frac{a_1}{b_1}, \frac{a_2}{b_2}, \ldots, \frac{a_n}{b_n}$, for $a_i, b_i > 0$, the sequence:
    $z_i = \frac{\sum_{j=i}^n a_j}{\sum_{j=i}^n b_j}$
    is also monotone increasing.
\end{lemma}

@@

\begin{lemma}
\label{lemma:any_ratio_is_sigma_simple}
For $j\ge 1$,
$$\frac{Pr[\bm{K_j}=\bm{k_j} \mid \bm{n_j}, H_a]}{Pr[\bm{K_j}=\bm{k_j} \mid \bm{n_j}, H_0]} = \sigma(k_j, p_a, p_0, n_j).$$
\end{lemma}
\begin{proof}
We induct on the number of rounds.
For $j=1$, we have
$$\frac{Pr[\bm{K_1}=\bm{k_1} \mid \bm{n_1},H_a]}{Pr[\bm{K_1}=\bm{k_1} \mid  \bm{n_1},H_0]} =\frac{Pr[K_1 = k_{1} \mid n_1,H_a]}{Pr[K_1 = k_1 \mid n_1,H_0]} $$$$= \frac{\text{Bin}(k_1,n_1,p_a)}{\text{Bin}(k_1,n_1,p_0)}=\sigma(k_1, p_a, p_0, n_1).$$
Suppose the lemma is true for round $j=m$ with history $\bm{k_m}$.
Observe that
 $$\frac{Pr[\bm{K_{m+1}}=\bm{k_{m+1}} \mid \bm{n_{m+1}},H_a]}{Pr[\bm{K_{m+1}}=\bm{k_{m+1}} \mid \bm{n_{m+1}}, H_0]} $$$$= \frac{ Pr[\bm{K_{m}}=\bm{k_{m}}\mid \bm{n_{m+1}},H_a] \cdot Pr[K_{m+1}'=k_{m+1}'|\bm{k_m},\bm{n_{m+1}},H_a]}{ Pr[\bm{K_{m}}= \bm{k_{m}} \mid  \bm{n_{m+1}},H_0]  \cdot  Pr[K_{m+1}'=k_{m+1}'|\bm{k_m},\bm{n_{m+1}},H_0]  }$$
 $$=\sigma(k_m, p_a, p_0, n_m) \cdot \frac{Pr[K_{m+1}'=k_{m+1}'|\bm{k_m}, \bm{n_{m+1}}, H_a]}{Pr[K_{m+1}'=k_{m+1}'|\bm{k_m},\bm{n_{m+1}},H_0]}$$
 by the induction hypothesis.
Then this is simply equal to
 $$\sigma(k_m, p_a, p_0, n_m)\cdot\frac{\text{Bin}(k_{m+1}',n_{m+1}',p_a)}{\text{Bin}(k_{m+1}',n_{m+1}',p_0)} = \sigma(k_{m+1}, p_a, p_0, n_{m+1})
 $$
 %$$
 %=\frac{p_a^{k_m} (1-p_a)^{n_m-k_m}}{p_0^{k_m} (1-p_0)^{n_m-k_m}} \cdot
% \frac{p_a^{k_{m+1}'} (1-p_a)^{n_{m+1}'-k_{m+1}'}}{p_0^{k_{m+1}'} (1-p_0)^{n_{m+1}'-k_{m+1}'}}
 %$$
 %$$
 %=\sigma(k_{m+1}, p_a, p_0, n_{m+1})
 %$$
\end{proof}

\begin{definition} 
\label{def:kmin}
Let $[n_1, \ldots, n_j]$ be the round schedule of an audit that has not stopped by the round $j-1$. Let us define 
\begin{small}
\begin{equation}\label{eq:kMin}
k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  =
  \min\left\{k : \omega_j(k, k_{j-1},p_a,p_0,n_j, n_{j-1}) \geq \frac{1}{\alpha}  \right\}.
%  \min\left\{k : \sigma(k_{r-1},p_a,p_0,n_{r-1})\cdot \tau_1(k-k_{r-1},p_a,p_0,n_r-n_{r-1}) \geq \frac{1}{\alpha}  \right\}$
\end{equation}
\end{small}
\end{definition}
As we have seen in Lemma \ref{lemma:minerva2_kmin_exists}, such a value of $k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}$ exists and $k_j \geq k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $ if and only if the result of the audit is Correct. 
% (\textit{i.e.,} the stopping condition in Definition~\ref{def:minervatwo} holds).

%The following lemma shows a Markov-like property of \Providence audit (\textit{i.e.,}
%for an audit that has not stopped in the first $j-1$ rounds, only cumulative results of the round $j-1$ matter: cumulative sample size $n_{j-1}$ and the number of ballots for the winner $k_{j-1}$).

%\textbf{Lemma \ref{lemma:markov}.}
%Let $[n_1, \ldots, n_{j-1}, n_j]$ be a round schedule for an execution of  \Providence audit that has not stopped
%in any of its first $j-1$ rounds (\textit{i.e.,} for every $i = 1, \ldots, j-1$:
%$k_i < k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} $), then: 
%\[ 
%$k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j} = k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j}$.
%\]
% \fpo{this can be used to prove that \Providence is more efficient than \Minerva and \BRAVO}
%\begin{proof}
%Let $k_{j-1}$ denote the number of ballots drawn for the declared winner up to the round $j-1$ (out of $n_{j-1}$ sampled ballots). The stopping decision for the round $j$ is made as follows:
%
%\[
% k^{p_a, p_0, \alpha, k_{j-1}}_{min, j, n_{j-1}, n_j}  = \min\left\{k : \omega_{j}(k, k_{r-1}, p_a, p_0, n_r, n_{r-1}) \geq \frac{1}{\alpha}  \right\} = 
%\]
%\[
%  =  k^{p_a, p_0, \alpha, k_{j-1}}_{min, 2, n_{j-1}, n_j}  
%\]
%This is easily observed. 
%\end{proof}

%That is, the stopping condition is equivalent to that of a two round audit with the same cumulative votes for the winner and cumulative round sizes: the first round is of size $n_{j-1}$ and has $k_{j-1}$ votes for the winner, and the second (cumulative) round size is $n_j$ with $k_j$ (cumulative) votes for the winner. Compare this to the similar property for the $\Bravo$ stopping condition.

### 2023-02-13 providence/sections/intro.tex
mostly proofs

\subsection{Adversarial Model for RLAs}
Detailed descriptions of best practices for post-election audits may be found in \cite{best-practices,why-and-how}. For our purposes, we will assume that the best practices are followed: the paper trail consists of hand-marked paper ballots and is secured; a public compliance audit is carried out before the RLA to ensure that the processes for securing the paper trail were followed; voter authentication and registration processes are verified, and only legitimate voters cast no more than a single vote each; the risk-limiting audit is public. We will further assume that all software used in the RLA is open source and well-defined, so its output may be reproduced and thus verified by an observer wishing to do so with their own software.

@@

We define a {\em weak adversary} as one who can choose the first and subsequent round sizes before the audit begins and a {\em strong adversary} as one who can choose any round size at any time before the round begins.

>>>>

\subsection{Adversarial Model for RLAs}
\label{sec:adv}
Detailed descriptions of best practices for post-election audits may be found in \cite{best-practices,why-and-how}. For our purposes, we will assume that the best practices are followed: the paper trail consists of hand-marked paper ballots and is secured; a public compliance audit is carried out before the RLA to ensure that the processes for securing the paper trail were followed; voter authentication and registration processes are verified, and only legitimate voters cast no more than a single vote each; the risk-limiting audit is public. We will further assume that all software used in the RLA is open source and well-defined, so its output may be reproduced and thus verified by an observer wishing to do so with their own software.

@@

We define a {\em weak adversary} as one who can choose the first and subsequent round sizes before the audit begins and a {\em strong adversary} as one who can choose any round size at any time (before, of course, that particular round begins). In particular, a strong adversary may use information about the sample drawn thus far to decide the next round size, but a weak adversary may not.

### 2023-02-13 providence/sections/prov.tex
mostly proofs

Recall that the proof that the \Minerva audit is risk-limiting assumes that the round schedule of \Minerva is predetermined and that, in particular, an adversarial auditor cannot determine the next round size after drawing a sample. This presents difficulties because a non-adversarial election official might want to draw a small next round if the current sample comes close to satisfying the risk limit. Because the \Minerva round size is predetermined, however, the election official would be required to draw a larger round size than necessary for the sample. Conversely, if the current sample is not at all close to satisfying the risk limit, it would be advantageous to draw a larger round than the predetermined round size.

We now formalize these notions of weak and strong adversaries.

\begin{definition}[Weak Adversary]
A {\emph weak adversary} may choose the first round size as a function of audit parameters. That is, the first round size is a function $$n_1(\alpha, p_a, p_0).$$
\end{definition}

\begin{definition}[Strong Adversary]
A {\emph strong adversary} may choose any round size as a function of audit parameters and all precreding samples. That is, the first round size is a function 
$$n_1(\alpha, p_a, p_0),$$
and for all rounds $j\ge 2$, the round size is a function
$$n_{j+1}(\alpha, p_a, p_0, \bm{k_{j-1}}, \bm{n_{j-1}}).$$
\end{definition}

Before defining \Providence, we give some intuition for how it is designed to avoid the problem of \Minerva.
In round $j$, for every possible value of $K_j$ for which the audit stops, a risk is incurred. To obtain the total risk for the round, one adds the risks corresponding to each value of $K_j$ for which the audit can stop, weighted by the probability of drawing that value of $K_j$. The stopping condition provides relationships among various quantities.

In \Minerva, the stopping condition relates the weighted average of the risks to the weighted average of the stopping probabilities over all values of $K_j$, for a given round size. Separate relationships between risk and stopping probability are not available for individual values of $K_j$. If the next round size depends on $K_j$, we do not have expressions relating the risks, and so \Minerva may be vulnerable to an adversary choosing round sizes \cite{usenix_minerva, arxiv_athena}.

In \Providence, we choose a stopping condition that applies separately to the risk and stopping probabilities for each value of $K_j$, avoiding the problem of \Minerva, and allowing for optimal round size choices, which depend on the drawn sample. 
The \Providence audit is risk-limiting even if an adversarial auditor determines round sizes after drawing the sample, and next round size computations may use knowledge of the current sample.

@@

\begin{theorem}
\label{thm:minerva2_is_rla_new}
An $(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence audit is an
$\alpha$-RLA.
\end{theorem}
\begin{proof}
Let $\mathcal{A}=(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence.
Let $\bm{n_j}$ be the cumulative round sizes used in this
audit, with corresponding cumulative tallies of
ballots for the reported winner $\bm{k_j}$.
For round $j=1$, by Definitions \ref{def:minervatwo}
and \ref{def:minerva_ratio}, we see that
the $\mathcal{A}=\text{Correct}$ (the audit stops) only when
$$
\tau_1(k_{1},p_a,p_0,n_1)\\
=\frac{Pr[K_{1} \geq k_{1} \mid H_a, n_1]}{Pr[K_{1} \geq k_{1} \mid H_0, n_1]}
\ge \frac{1}{\alpha}.
$$
By Lemma \ref{lemma:minerva2_kmin_exists} and Definition \ref{def:kmin}, there is 
a value $k_{min,1} = k^{p_a, p_0, \alpha, 0}_{min, 1, 0, n_1}$ such that 
$$
\frac{Pr[K_{1} \geq k_{1} \mid H_a, n_1]}{Pr[K_{1} \geq k_{1} \mid H_0, n_1]}
\ge
\frac{Pr[K_{1} \geq k_{min,1} \mid H_a, n_1]}{Pr[K_{1} \geq k_{min, 1} \mid H_0, n_1]}
\ge 
\frac{1}{\alpha}.
$$

@@

Summing both sides over all values of $k_{j-1} < k_{min, j-1}$ gives us a similar relationship between the probabilities of stopping in round $j$ (given the null and alternate hypotheses respectively).
When both sides of the inequality are further summed over all rounds, we get:

>>>>

Recall that the proof that the \Minerva audit is risk-limiting assumes that the round schedule of \Minerva is predetermined and that, in particular, an adversarial auditor cannot determine the next round size after drawing a sample (we termed such an adversary a weak adversary in section \ref{sec:adv}). This presents difficulties because a non-adversarial election official might want to draw a small next round if the current sample comes close to satisfying the risk limit. Because the \Minerva round size is predetermined, however, the election official would be required to draw a larger round size than necessary for the sample. Conversely, if the current sample is not at all close to satisfying the risk limit, it would be advantageous to draw a larger round than the predetermined round size. Thus it would be advantageous to have an RLA with \Minerva's efficiency advantages in the face of an adversarial auditor who could choose next round sizes after drawing the sample (we termed such an adversary a strong adversary). This would enable honest participants to better plan audits.

Before proceeding to define the audit \Providence which is an RLA in the presence of the stronger adversary, we formalize the notions of weak and strong adversaries introduced in section \ref{sec:adv}.

\begin{definition}[Weak Adversary]
A {\emph weak adversary} may choose the first and consequent round sizes as a pre-determined function of audit parameters. That is, the $j^{th}$ round size is a function $$n_j(\alpha, p_a, p_0)$$ determined before the audit begins. 
\end{definition}

\begin{definition}[Strong Adversary]
A {\emph strong adversary} may choose any round size as any function of audit parameters and all preceding samples. That is, the first round size is a function 
$$n_1(\alpha, p_a, p_0),$$
and for all rounds $j\ge 2$, the round size is a function
$$n_{j}(\alpha, p_a, p_0, \bm{k_{j-1}}, \bm{n_{j-1}})$$ The functions $n_j$, $j \geq 1$ may be chosen at any time before the $j^{th}$ round begins. 
\end{definition}

Before defining \Providence, we give some intuition for how it is designed to avoid the problem of \Minerva.

For any ballot polling audit, in round $j$, designed to stop for some value of $K_j$, however large $K_j$ is, there is, in general, a probability, however small, that the value of $K_j$ has resulted from an underlying incorrect election outcome. Thus, corresponding to each such value of $K_j$, a risk---probability that the audit will stop given that the outcome is incorrect---is incurred. To obtain the total risk for the round, one adds the risks corresponding to each value of $K_j$ for which the audit can stop, weighted by the probability of drawing that value of $K_j$. The stopping condition may provide relationships among the various quantities.

In \Minerva, the stopping condition relates the weighted average of the risks to the weighted average of the stopping probabilities over all values of $K_j$, for a given round size. Separate relationships between risk and stopping probability are not available for individual values of $K_j$. If the next round size depends on $K_j$, expressions relating the risks are not available, and so we are not able to say if \Minerva is an RLA in this case, or if it is vulnerable to an adversary choosing round sizes \cite{usenix_minerva, arxiv_athena}.

In \Providence, we choose a stopping condition that applies separately to the risk and stopping probabilities for each value of $K_j$, avoiding the problem of \Minerva, and allowing for optimal round size choices, which depend on the drawn sample. 
The \Providence audit is risk-limiting even if a strong adversarial auditor determines round sizes after drawing the sample, and next round size computations may use knowledge of the current sample.

@@

\begin{theorem}
\label{thm:minerva2_is_rla_new}
$(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence is an
$\alpha$-RLA.
\end{theorem}
\begin{proof}
Let $\mathcal{A}=(\alpha,p_a, p_0,k_{j-1},n_{j-1},n_j)$-\Providence.
Let $\bm{n_j}$ be the cumulative round sizes used in this
audit, with corresponding cumulative tallies of
ballots for the reported winner $\bm{k_j}$.
For round $j=1$, by Definitions \ref{def:minervatwo}
and \ref{def:minerva_ratio}, we see that
the $\mathcal{A}=\text{Correct}$ (the audit stops) only when
$$
\tau_1(k_{1},p_a,p_0,n_1)\\
=\frac{Pr[K_{1} \geq k_{1} \mid H_a, n_1]}{Pr[K_{1} \geq k_{1} \mid H_0, n_1]}
\ge \frac{1}{\alpha}.
$$
By Lemma \ref{lemma:minerva2_kmin_exists} and Definition \ref{def:kmin}, there is 
a value $k_{min,1} = k^{p_a, p_0, \alpha, 0}_{min, 1, 0, n_1}$ such that 
$$
\frac{Pr[K_{1} \geq k_{1} \mid H_a, n_1]}{Pr[K_{1} \geq k_{1} \mid H_0, n_1]}
\ge
\frac{Pr[K_{1} \geq k_{min,1} \mid H_a, n_1]}{Pr[K_{1} \geq k_{min, 1} \mid H_0, n_1]}
\ge 
\frac{1}{\alpha}.
$$

@@

Summing both sides over all values of $k_{j-1} < k_{min, j-1}$ gives us a similar relationship between the probabilities of stopping in round $j$ (given the null and alternate hypotheses respectively)\footnote{In fact, this is the relationship \Minerva enforces for its stopping condition, additionally requiring that $n_j$ be fixed for all values of $k_{j-1}$.}. 
When both sides of the inequality are further summed over all rounds, we get:
